The brightness and spatial distributions of terrestrial radio sources

### 2013-07-08 articles/Article-RFI-distribution/article-rfi-distribution.tex
Ready for resubmit\!

\newcommand{\editmark}[1]{{\color{red}{\textbf{#1}}}}

@@

Therefore, if we plot the histogram of the RFI amplitudes in a $\log$-$\log$ plot, we expect to see a power law with a slope of $-2$ over the interval in which the RFI sources are spread like uniform sources on a two-dimensional plane. For this, we have assumed that the variables $I$, $g$ and $\rho$ are constants. In reality, these variables have a stochastic nature. The effect of this will be discussed in \S\ref{sec:parameter-variability}.

@@

\subsection{Propagation effects}
So far, we have assumed that the electromagnetic radiation propagates through free space, resulting in an $r^{-2}$ fall-off. In reality, the radiation will be affected by complicated propagational effects \editmark{due to the surface of the Earth}. %Because of the irregular surface of the Earth (including urban areas) and the absence of a direct line of sight between transmitter and receiver, the propagation mode might be indirect. Multiple indirect paths might be formed that include reflection, refraction, diffraction and absorption of the electromagnetic wave. 
A commonly used propagation model is the empirical model determined by \citet{okumura-propagation-model}, which was further developed by \citet{hata-propagation-loss}. The original model is based on urban areas, but corrections are given by Hata for suburban areas with a lower population density and for open areas. Hata gives the following analytical estimate for $L_p$, the electromagnetic propagation loss between two ground-based antennas:
\begin{align} 
\notag                L_p & = & 69.55 + 26.16 \log_{10} f_c - 13.82 \log_{10} h_b - \\
\label{eq:hata-model} & &  a(h_m) + (44.9 - 6.55 \log_{10} h_b) \log_{10} r, 
\end{align}
where $L_p$ the loss in dB; $f_c$ the radiation frequency in MHz; $h_b$ the height of the transmitting antenna in meters; $h_m$ the height of the receiving antenna in meters; $r$ the distance between the antennas in meters; and $a(h_m)$ a correction factor in dB that corrects for the height of the receiving antenna and the urban density. \citeauthor{hata-propagation-loss} found this model to be representative for frequencies $f_c \sim$~150--1500~MHz, with transmitter heights $h_b \sim$~30--200~m, receiver heights $h_m \sim$~1--10~m and over distances $r\sim$~1--20~km.

@@

@@

@@

\section{Conclusions and discussion} \label{sec:dist-discussion}
We have analysed the histogram of visibility amplitudes of LOFAR observations and found that, within a significant range of the histogram, the contribution of RFI sources follows a power-law distribution. The found power-law exponents of $-1.62$ and $-1.53$ for the 30--78~MHz LBA and 115--163~MHz HBA observations respectively, can be explained by a uniform spatial distribution of RFI sources, affected by propagation described surprisingly well by Hata's electromagnetic propagation model. Taken at face value these exponents imply in Hata's model that the average transmitting heights for sources affecting the LBA and HBA are $79$ and $13$~m respectively. There are no 79~m high transmitters nearby LOFAR stations in the LBA frequency range. Additionally, Hata's model only goes down to 150~MHz, and it is possible that the electromagnetic fall-off due to propagation will be different for lower frequencies, e.g. because the effect of the ionosphere becomes stronger. Intervals for the exponents with representative $3\sigma$ boundaries are $[-1.80;-1.44]$ for the LBA and $[-1.57;-1.49]$ for the HBA. This translates to boundaries on the average transmitter height of $[0.6; 800]$ and $[3.1; 23]$~m for the LBA and HBA respectively. Therefore, the LBA measurements are clearly not accurate enough to be conclusive. Moreover, because the power-law distribution analyses involve many assumptions, it is uncertain whether the analyses are sufficiently accurate for making these detailed conclusions.

@@

\subsection{Implications for very long integrations}
In theory, faint RFI could impose a fundamental limit on the attainable noise limit of long integrations. \editmark{We} will analyse the situation for the LOFAR EoR project. This project will use the LOFAR high-band antennas to collect on the order of 50--100 night-time observations of 6 h for a few target fields. The final resolution required for signal extraction will be about $1$~MHz. The project will use about $60$~stations, each of which provides two polarized feeds. This will bring the noise level in a single 6~h observation in 1~MHz bandwidth to
\begin{equation}
 \sigma_\textrm{eor-night} = \textrm{SEFD} \left(2\Delta t \Delta \nu N_\textrm{feed} N_\textrm{interferometers} \right)^{-\frac{1}{2}} \approx 250\hspace{1mm} \mu\textrm{Jy},
\end{equation}
where $N_\textrm{feed}=2$ is the number of feeds per antenna and $N_\textrm{interferometers}=\frac{1}{2}60\times59$ is the number of interferometers. Therefore, after 100 nights the thermal noise level will be $25\hspace{1mm}\mu\textrm{Jy}$.

Because some RFI sources might be stationary, the signals from these sources will add consistently over time, meaning that the geometrical phase will be the same every day. Therefore, the amount that time integration can decrease the flux density of RFI might be limited. %Additionally, some RFI sources are received by multiple stations of the array, and by multiple feeds of the individual antennas. Therefore, integrating data from different interferometers and data from the two polarized feeds might not lower the noise level that is caused by RFI at some point. In summary, RFI is unlike normal noise and might add consistent over time, interferometer and feed.
\editmark{On the other hand}, many RFI signals observed in the LOFAR bands have a limited bandwidth. Indeed, the majority of the detected RFI sources affect only one or a few LOFAR channels of 0.76 kHz. Therefore, frequency averaging will lower the flux density of the RFI signal. If the frequency range contains only one stationary RFI source, the strength of this source will go down linearly with the total bandwidth. If we assume that all channels are affected by RFI sources and all these sources transmit in approximately one channel, then the noise addition that is produced by RFI will go down with the square root of the number of averaged channels. This is a consequence of the random phase that different RFI sources have.

@@

\subsection{Interference-reducing effects} \label{sec:coherence-reduction}
When integrating data, it is likely that the actual noise limit from low-level RFI will be significantly lower than the given upper limit, which was determined at highest LOFAR resolution. \editmark{There are several reasons for this: Many RFI sources have a variable geometric phase, because they move or because their path of propagation changes; many RFI sources will be seen by only a few stations or are not constant over time; for the shortest baselines at 150~MHz, the far field starts around 1~km, and some RFI sources will therefore be in the near field; and finally, a large number of stationary RFI sources in a uniform spatial distribution will interfere both constructively and destructively with each other.}

\editmark{Fringe} stopping interferometers can partly average out RFI sources. Nevertheless, stationary RFI that is averaged out by fringe stopping will leave artefacts in the field centre \citep{post-correlation-filtering}. This is not relevant when observing the North Celestial Pole --- which is one of the LOFAR EoR fields --- because no fringe stopping is applied when observing the NCP. \editmark{Imaging of the interferometric data} will localize the contribution from stationary RFI near the NCP. If RFI artefacts would show in the image of the NCP field, they can easily be detected and possibly be removed, or processing could ignore data near the pole. Because of these arguments, when considering RFI it is a risk to use the NCP as one of the EoR target fields. At the same time, this field is useful for analysing the RFI coherency properties. Preliminary analysis of EoR NCP observations of a single night have almost reached the thermal noise, but do not show leaked RFI at the pole \citep[\S4.3]{ncp-eor-yatawatta}.

\editmark{Our given RFI constraints could be too pessimistic, because} we have assumed 100\% of the spectrum is occupied by RFI over the entire duration of the observation. If, say, only 10\% of the samples are affected by RFI, the expected value of the leaked RFI level decreases by about $25$\%, and if the detected 2.68\% true-positives contain all RFI, there is no leaked RFI at all. With current data, one can only speculate how much of the electromagnetic spectrum is truly occupied.

>>>>

%\newcommand{\editmark}[1]{#1}
\newcommand{\editmark}[1]{{\color{red}{\textbf{#1}}}}

@@

Therefore, if we plot the histogram of the RFI amplitudes in a $\log$-$\log$ plot, we expect to see a power law with a slope of $-2$ over the interval in which the RFI sources are spread like uniform sources on a two-dimensional \editmark{plane.} %For this, we have assumed that the variables $I$, $g$ and $\rho$ are constants. In reality, these variables have a stochastic nature. The effect of this will be discussed in \S\ref{sec:parameter-variability}.

@@

\subsection{Propagation effects}
So far, we have assumed that the electromagnetic radiation propagates through free space, resulting in an $r^{-2}$ fall-off. In reality, the radiation will be affected by complicated propagational effects \editmark{due to the surface of the Earth}. %Because of the irregular surface of the Earth (including urban areas) and the absence of a direct line of sight between transmitter and receiver, the propagation mode might be indirect. Multiple indirect paths might be formed that include reflection, refraction, diffraction and absorption of the electromagnetic wave. 
A commonly used propagation model is the empirical model determined by \citet{okumura-propagation-model}, which was further developed by \citet{hata-propagation-loss}. %The original model is based on urban areas, but corrections are given by Hata for suburban areas with a lower population density and for open areas.
\editmark{Hata} gives the following analytical estimate for $L_p$, the electromagnetic propagation loss between two ground-based antennas:
\begin{align} 
\notag                L_p & = & 69.55 + 26.16 \log_{10} f_c - 13.82 \log_{10} h_b - \\
\label{eq:hata-model} & &  a(h_m) + (44.9 - 6.55 \log_{10} h_b) \log_{10} r, 
\end{align}
where $L_p$ the loss in dB; $f_c$ the radiation frequency in MHz; $h_b$ the height of the transmitting antenna in meters; $h_m$ the height of the receiving antenna in meters; $r$ the distance between the antennas in meters; and $a(h_m)$ a correction factor in dB that corrects for the height of the receiving antenna and the urban density. \citeauthor{hata-propagation-loss} found this model to be representative for frequencies $f_c \sim$~150--1500~MHz, with transmitter heights $h_b \sim$~30--200~m, receiver heights $h_m \sim$~1--10~m and over distances $r\sim$~1--20~km.

@@

\editmark{We note that a uniform distribution of meteors or aircrafts which reflect free-space propagating RFI can create a power-law distribution with a similar slope: a uniform two-dimensional distribution of reflecting sources will create a $-1.5$ slope, while a uniform three-dimensional distribution will create a $-1.75$ slope. With brightness-distribution analyses one can therefore not distinguish between transmitters affected by Hata's propagation model and reflectors affected by free-space propagation. Reflected RFI might become relevant at lower amplitude levels.}

@@

\editmark{From Eq.~\eqref{eq:frequency-density-with-propagation} it can be seen that the $\rho$, $I$ and $g$ parameters have the same effect of scaling the power-law distribution, and do not change its shape or slope. Therefore, with distribution analyses one can e.g. not determine whether the distribution is dominated by low-power sources within the horizon or by scattered signals from over the horizon. The horizon of an antenna is estimated 	with $\sqrt{2hr}$ \citep{bullington-propagation}, with $r$ the radius of the Earth and $h$ the height of the antenna. For LOFAR, the horizon is at about 5~km.}

@@

\section{Conclusions and discussion} \label{sec:dist-discussion}
We have analysed the histogram of visibility amplitudes of LOFAR observations and found that, within a significant range of the histogram, the contribution of RFI sources follows a power-law distribution. The found power-law exponents of $-1.62$ and $-1.53$ for the 30--78~MHz LBA and 115--163~MHz HBA observations respectively, can be explained by a uniform spatial distribution of RFI sources, affected by propagation described surprisingly well by Hata's electromagnetic propagation model. Taken at face value these exponents imply in Hata's model that the average transmitting heights for sources affecting the LBA and HBA are $79$ and $13$~m respectively. There are no 79~m high transmitters nearby LOFAR stations in the LBA frequency range. Additionally, Hata's model only goes down to 150~MHz, and it is possible that the electromagnetic fall-off due to propagation will be different for lower frequencies. Intervals for the exponents with representative $3\sigma$ boundaries are $[-1.80;-1.44]$ for the LBA and $[-1.57;-1.49]$ for the HBA\editmark{, giving} average transmitter heights of $[0.6; 800]$ and $[3.1; 23]$~m for the LBA and HBA respectively. Therefore, the LBA measurements are clearly not accurate enough to be conclusive. Moreover, because the power-law distribution analyses involve many assumptions, it is uncertain whether the analyses are sufficiently accurate for making these detailed conclusions.

@@

\subsection{Implications for very long integrations}
Faint RFI could impose a fundamental limit on the attainable noise limit of long integrations. \editmark{We} will analyse the situation for the LOFAR EoR project. This project will use the LOFAR high-band antennas to collect on the order of 50--100 night-time observations of 6 h for a few target fields. The final resolution required for signal extraction will be about $1$~MHz. The project will use about $60$~stations, each of which provides two polarized feeds. This will bring the noise level in a single 6~h observation in 1~MHz bandwidth to
\begin{equation}
 \sigma_\textrm{eor-night} = \textrm{SEFD} \left(2\Delta t \Delta \nu N_\textrm{feed} N_\textrm{interferometers} \right)^{-\frac{1}{2}} \approx 250\hspace{1mm} \mu\textrm{Jy},
\end{equation}
where $N_\textrm{feed}=2$ is the number of feeds per antenna and $N_\textrm{interferometers}=\frac{1}{2}60\times59$ is the number of interferometers. Therefore, after 100 nights the thermal noise level will be $25\hspace{1mm}\mu\textrm{Jy}$.

Because some RFI sources might be stationary, the signals from these sources will add consistently over time, meaning that the geometrical phase will be the same every day. Therefore, the amount that time integration can decrease the flux density of RFI might be limited. %Additionally, some RFI sources are received by multiple stations of the array, and by multiple feeds of the individual antennas. Therefore, integrating data from different interferometers and data from the two polarized feeds might not lower the noise level that is caused by RFI at some point. In summary, RFI is unlike normal noise and might add consistent over time, interferometer and feed.
\editmark{On the other hand}, many RFI signals observed in the LOFAR bands have a limited bandwidth, and the majority of the detected RFI sources affect only one or a few LOFAR channels of 0.76 kHz. Therefore, frequency averaging will lower the flux density of the RFI signal. If the frequency range contains only one stationary RFI source, the strength of this source will go down linearly with the total bandwidth. If we assume that all channels are affected by RFI sources and all these sources transmit in approximately one channel, then the noise addition that is produced by RFI will go down with the square root of the number of averaged channels. This is a consequence of the random phase that different RFI sources have.

@@

\subsection{Interference-reducing effects} \label{sec:coherence-reduction}
When integrating data, it is likely that the actual noise limit from low-level RFI will be significantly lower than the given upper limit, which was determined at highest LOFAR resolution. \editmark{There are several reasons for this: Many RFI sources have a variable geometric phase, because they move or because their path of propagation changes; many RFI sources will be seen by only a few stations or are not constant over time; for the shortest baselines at 150~MHz, the far field starts around 1~km, and some RFI sources will therefore be in the near field; and finally, a large number of stationary RFI sources in a uniform spatial distribution will interfere both constructively and destructively with each other. These arguments are valid only for interferometric arrays. Global EoR experiments that use a single antenna will not benefit from these effects, and will still be limited by low-level RFI.}

\editmark{Fringe} stopping interferometers can partly average out RFI sources. Nevertheless, stationary RFI that is averaged out by fringe stopping will leave artefacts in the field centre \citep{post-correlation-filtering}. This is not relevant when observing the North Celestial Pole --- which is one of the LOFAR EoR fields --- because no fringe stopping is applied when observing the NCP. \editmark{Imaging of the data} will localize the contribution from stationary RFI near the NCP. If RFI artefacts would show in the image of the NCP field, they can easily be detected and possibly be removed, or processing could ignore data near the pole. Because of these arguments, it is a risk to use the NCP as one of the EoR target fields. At the same time, this field is useful for analysing the RFI coherency properties. Preliminary analysis of EoR NCP observations of a single night have almost reached the thermal noise, but do not show leaked RFI at the pole \citep[\S4.3]{ncp-eor-yatawatta}.

\editmark{Because we have assumed 100\% of the spectrum is occupied by RFI, our given RFI constraints could be too pessimistic. If} only 10\% of the samples are affected by RFI, the expected value of the leaked RFI level decreases by about $25$\%, and if the detected 2.68\% true-positives contain all RFI, there is no leaked RFI at all. With current data, one can only speculate how much of the electromagnetic spectrum is truly occupied.
